The Evolution of Phenotypically Invariant Genetic Networks

### 2017-12-08 kalman_walk_paper.tex
removed other apx

\subsection*{Hybrid incompatibility}
Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
if the parents have different underlying systems.
%This leads to the question: 
How quickly do hybrid phenotypes break down as genetic distance between parents increases?
We will quantify how far a system's phenotype is from optimal
using a weighted difference between impulse response functions.
Suppose that $\rho(t)$ is a nonnegative, smooth, square-integrable weighting function,
$h_0(t)$ is the \emph{optimal} impulse response function
and define the ``distance to optimum'' of another impulse response function
to be
\begin{align}
\label{eqn:distance}
	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
\end{align}
Consider reproduction between a parent with system $(A, B, C)$ 
and another displaced by distance $\epsilon$ in the direction $(X,Y,Z)$,
i.e., having  system $(A + \epsilon X, B + \epsilon Y, C + \epsilon Z)$.
We assume both are ``perfectly adapted'' systems, 
i.e., having impulse response function $h_0(t)$,
and their offspring has impulse response function $h_\epsilon(t)$.
A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ is explicitly worked out in Appendix \ref{apx:H_calc},
and shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
while $F_2$ hybrids are at distance proportional to $\epsilon$.
This is because an $F_1$ hybrid has one copy of each parental system,
and therefore lies directly between the parental systems (see Figure \ref{fig:conceptual_fig}) --
the parents both lie in $\allS$, which is the valley defined by $D$,
and so their midpoint only differs from optimal due to curvature of $\allS$.
In contrast, an $F_2$ hybrid may be homozygous for one parental type in some coefficients
and homozygous for the other parental type in others;
this means that each coefficient of an $F_2$ may be equal to either one of the parents,
or intermediate between the two;
this means that possible $F_2$ systems may be as far from the optimal set, $\allS$,
as the distance between the parents.
The precise rate at which the phenotype of a hybrid diverges depends on the geometry
of the optimal set $\allS$ relative to segregating genetic variation.
% in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set) with respect to the coordinates.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Local expansion of the fitness surface}
\label{apx:H_calc}

Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
With $h_0(t) = C_0 e^{tA_0} B_0$ a representative of the optimal set:
\begin{equation}
    \begin{aligned}
        D(A, B, C)^2
        &:= 
        \int_0^\infty \rho(t) \left| h_A(t) - h_0(t) \right|^2 dt \\
        &:= 
        \int_0^\infty \rho(t) \left| C e^{At} B - C_0 e^{A_0 t} B_0 \right|^2 dt \\
        &= 
        \int_0^\infty \rho(t) \tr\left\{
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
        \right\} dt \\
        &= 
        \int_0^\infty \rho(t) \tr\left\{
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
        \right\} dt  ,
        % &= 
        % \int_0^\infty \rho(t)  \tr\left\{
        %         C e^{At} B B^T e^{A^T t} C^T 
        %         - C_0 e^{A_0t} B_0 B^T e^{A^T t} C^T 
        %         \right. \\ &\qquad \qquad \left. {}
        %         - C e^{At} B B_0^T e^{A_0^T t} C_0^T 
        %         + C_0 e^{A_0t} B_0 B_0^T e^{A_0^T t} C_0^T 
        % \right\} dt \\
        % \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
        % &= 
        % \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
    \end{aligned}
\end{equation}
where $\tr X$ denotes the trace of a square matrix $X$.
How does this change as we perturb about $(A_0, B_0, C_0)$?
First we differentiate with respect to $A$, keeping $B=B_0$ and $C=C_0$ fixed.
Since
\begin{equation}
  \begin{aligned}
      \frac{d}{du} e^{(A+uZ)t} \vert_{u=0}
      &=
      \int_0^t e^{As} Z e^{A(t-s)} ds, 
  \end{aligned}
\end{equation}
we have that
\begin{equation}
  \begin{aligned}
      \frac{d}{du} D(A+uZ,B_0,C_0)^2 \vert_{u=0}
      &=
        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 B_0^T \left( e^{At} - e^{A_0 t} \right)^T C_0^T \right\} dt \\
      &=
        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 \left( h_A(t) - h_0(t) \right)^T \right\} dt 
  \end{aligned}
\end{equation}
and, by differentiating this and supposing that $A$ is on the optimal set,
i.e., $h_A(t)=h_0(t)$, (so without loss of generality, $A=A_0$):
\begin{equation}
  \begin{aligned}
      \calH^{A,A}(Y,Z) 
      &:= 
      \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ,B_0,C_0)^2 \vert_{u=v=0} \\
      &=
        \int_0^\infty \rho(t) \tr\left\{ C_0 
        \left( \int_0^t e^{A_0 s} Y e^{A_0 (t-s)} ds \right) 
        B_0 B_0^T 
        \left( \int_0^t e^{A_0 s} Z e^{A_0 (t-s)} ds \right)^T
        C_0^T \right\} dt  .
  \end{aligned}
\end{equation}

The function $\calH$ will define a quadratic form.
To illustrate the use of this, suppose that $B$ and $C$ are fixed.
% Here $\calH$ is the quadratic form underlying the Hamiltonian.
By defining $\Delta_{ij}$ to be the matrix with a 1 in the $(i,j)$th slot
and 0 elsewhere,
the coefficients of the quadratic form is
\begin{equation}
    \begin{aligned}
        H_{ij, k\ell}(A)
        &:=
        \calH(\Delta_{ij}, \Delta_{k\ell}) .
    \end{aligned}
\end{equation}

We could use this to get the quadratic approximation to $D$ near the optimal set.
To do so, it'd be nice to have a way to compute the inner integral above.
Suppose that we diagonalize $A = U \Lambda U^{-1}$.
Then
\begin{equation} \label{eqn:exp_deriv}
  \begin{aligned}
      \int_0^t e^{As} Z e^{A(t-s)} ds 
      &=
      \int_0^t U e^{\Lambda s} U^{-1} Z U e^{\Lambda (t-s)} U^{-1} ds 
  \end{aligned}
\end{equation}
Now, notice that
\begin{equation}
  \begin{aligned}
      \int_0^t e^{s \lambda_i} e^{(t-s) \lambda_j} ds
      &=
      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
          \qquad & \text{if} \quad i \neq j \\
      &=
          t e^{t \lambda_i} 
          \qquad & \text{if} \quad i = j 
  \end{aligned}
\end{equation}
Therefore, 
defining
\begin{equation}
    \begin{aligned}
    X_{ij}(t,Z) 
       &= 
        \left( U^{-1} Z U \right)_{ij}
      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
          \qquad & \text{if} \quad i \neq j \\
      &=
          \left( U^{-1} Z U \right)_{ii}
          t e^{t \lambda_i} 
          \qquad & \text{if} \quad i = j 
    \end{aligned}
\end{equation}
moving the $U$ and $U^{-1}$ outside the integral and integrating we get that
\begin{equation}
  \begin{aligned}
      \int_0^t e^{As} Z e^{A(t-s)} ds 
      &=
      U X(t,Z) U^{-1} .
  \end{aligned}
\end{equation}
This implies that
\begin{equation}
    \begin{aligned}
        D(A_0+\epsilon Z)^2
        &\approx \frac{1}{2} \epsilon^2 
        \int_0^\infty
            \rho(t) \tr\left\{ C U X(t,Z) U^{-1} B B^T (U^{-1})^T X(t,Z)^T U^T C^T \right\}
        dt .
    \end{aligned}
\end{equation}

To compute the $n^2 \times n^2$ matrix $H$,
we see that if $Z=\Delta_{k \ell}$, then
\begin{equation}
  \begin{aligned}
      X_{ij}^{k\ell}(t) 
      &= 
      (U^{-1})_{\cdot k} U_{\ell \cdot}
      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
          \qquad & \text{if} \quad i \neq j \\
      &=
      (U^{-1})_{\cdot k} U_{\ell \cdot}
      t e^{t \lambda_i} 
          \qquad & \text{if} \quad i = j 
  \end{aligned}
\end{equation}
where $U_{k \cdot}$ is the $k$th row of $U$,
and so
\begin{equation}
    \begin{aligned}
        H_{ij, k\ell}(A)
        &=
        \int_0^\infty
            \rho(t) \tr\left\{ C U X^{ij}(t) U^{-1} B B^T (U^{-1})^T X^{k\ell}(t)^T U^T C^T \right\}
        dt .
    \end{aligned}
\end{equation}
This implies that
\begin{equation}
    \begin{aligned}
        D(A_0+\epsilon Z)^2
        &\approx \frac{1}{2} \epsilon^2 \sum_{ijk\ell} H_{ij,k\ell}(A_0) Z_{ij} Z_{k\ell}  .
    \end{aligned}
\end{equation}

By section \ref{ss:quant_gen},
if we set $\Sigma=\sigma^2 I$ and $U=H$,
then a population at $A_0+Z$ experiences a restoring force of strength
$(I + \sigma^2 H^{-1})^{-1} Z$ (treating $Z$ as a vector and $H$ as an operator on these).
If $\sigma^2$ is small compared to $H^{-1}$
then this is approximately $-\sigma^2 H^{-1} Z$.
This suggests that the population mean follows an Ornstein-Uhlenbeck process,
as described (in different terms) in \citet{hansen1996translating}.

More generally, $B$ and $C$ may also change.
To extend this we need the remaining second derivatives of $D^2$.
First, in $B$:
\begin{equation}
    \begin{aligned}
        \mathcal{H}^{B,B}(Y,Z)
        &:= 
        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY+vZ,C_0)\vert_{u=v=0} \\
        &=
        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
        C_0 e^{t A_0}
        \frac{d}{du} \frac{d}{dv} 
        (uY+vZ)
        (uY+vZ)^T
        \vert_{u=v=0} 
        e^{t A_0^T} C_0^T 
        \right\} dt  \\
        &=
        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
        C_0 e^{t A_0}
        \left( Y Z^T + Z Y^T \right)
        e^{t A_0^T} C_0^T 
        \right\} dt  .
  \end{aligned}
\end{equation}
Next, in $C$:
\begin{equation}
    \begin{aligned}
        \mathcal{H}^{B,B}(Y,Z)
        &:= 
        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0,C_0+uY+vZ)\vert_{u=v=0} \\
        &=
        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
        B_0 e^{t A_0^T}
        \frac{d}{du} \frac{d}{dv} 
        (uY+vZ)^T
        (uY+vZ)
        \vert_{u=v=0} 
        e^{t A_0} B_0
        \right\} dt  \\
        &=
        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
        B_0 e^{t A_0^T}
        \left( Y Z^T + Z Y^T \right)
        e^{t A_0} B_0
        \right\} dt  .
  \end{aligned}
\end{equation}
Now, the mixed derivatives in $B$ and $C$:
\begin{equation}
    \begin{aligned}
        \mathcal{H}^{B,C}(Y,Z)
        &:= 
        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY,C_0+vZ)\vert_{u=v=0} \\
        &=
        \int_0^\infty \rho(t) \tr\left\{ 
        Y e^{t A_0^T} C_0^T Z e^{t A_0} B_0
        \right\} dt  .
  \end{aligned}
\end{equation}
In $A$ and $B$
\begin{equation}
    \begin{aligned}
        \mathcal{H}^{A,B}(Y,Z)
        &:= 
        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0+vZ,C_0)\vert_{u=v=0} \\
        &=
        \int_0^\infty \rho(t) \tr\left\{ 
        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
        Z^T e^{t A_0} C_0
        \right\} dt  ,
  \end{aligned}
\end{equation}
and finally in $A$ and $C$:
\begin{equation}
    \begin{aligned}
        \mathcal{H}^{A,C}(Y,Z)
        &:= 
        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0,C_0+vZ)\vert_{u=v=0} \\
        &=
        \int_0^\infty \rho(t) \tr\left\{ 
        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
        B_0 e^{t A_0} Z
        \right\} dt  .
  \end{aligned}
\end{equation}

Together, numerical computation of these expressions,
along with estimates of genetic covariance within a population,
allow precise predictions of evolutionary dynamics of a particular system.
The approximation should be good as long as the second-order Taylor approximation holds.

>>>>

\subsection*{Hybrid incompatibility}
Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
if the parents have different underlying systems.
%This leads to the question: 
How quickly do hybrid phenotypes break down as genetic distance between parents increases?
We will quantify how far a system's phenotype is from optimal
using a weighted difference between impulse response functions.
Suppose that $\rho(t)$ is a nonnegative, smooth, square-integrable weighting function,
$h_0(t)$ is the \emph{optimal} impulse response function
and define the ``distance to optimum'' of another impulse response function
to be
\begin{align}
\label{eqn:distance}
	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
\end{align}
Consider reproduction between a parent with system $(A, B, C)$ 
and another displaced by distance $\epsilon$ in the direction $(X,Y,Z)$,
i.e., having  system $(A + \epsilon X, B + \epsilon Y, C + \epsilon Z)$.
We assume both are ``perfectly adapted'' systems, 
i.e., having impulse response function $h_0(t)$,
and their offspring has impulse response function $h_\epsilon(t)$.
A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ % is explicitly worked out in Appendix \ref{apx:H_calc}, and
shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
while $F_2$ hybrids are at distance proportional to $\epsilon$.
This is because an $F_1$ hybrid has one copy of each parental system,
and therefore lies directly between the parental systems (see Figure \ref{fig:conceptual_fig}) --
the parents both lie in $\allS$, which is the valley defined by $D$,
and so their midpoint only differs from optimal due to curvature of $\allS$.
In contrast, an $F_2$ hybrid may be homozygous for one parental type in some coefficients
and homozygous for the other parental type in others;
this means that each coefficient of an $F_2$ may be equal to either one of the parents,
or intermediate between the two;
this means that possible $F_2$ systems may be as far from the optimal set, $\allS$,
as the distance between the parents.
The precise rate at which the phenotype of a hybrid diverges depends on the geometry
of the optimal set $\allS$ relative to segregating genetic variation.
% in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set) with respect to the coordinates.

@@
